Effective multi-body SU($N$)-symmetric interactions of ultracold fermionic atoms on a 3-D lattice

### 2019-03-18 multi_body_interactions/papers/theory/paper.tex
final fixes/revisions to multi-body theory paper

% "left vector" arrow; requires tikz package
\newcommand{\lvec}[1]
{\reflectbox{\ensuremath{\vec{\reflectbox{\ensuremath{#1}}}}}}

@@

While an external trapping potential will generally break discrete
translational symmetry of the lattice, any background inhomogeneity
can be made negligible by spectroscopically addressing a sufficiently
small region of the lattice \cite{goban2018emergence}.  Throughout
this paper, we work strictly in the deep-lattice regime with
negligible tunneling between lattice sites.  We also neglect any
lattice inhomogeneities and assume that both atomic orbital states
(i.e.~${}^1S_0$ and ${}^3P_0$) experience identical lattice
potentials.  The single-particle Hamiltonian of the atoms can then be
written in the form
\begin{align}
  H_0 = \sum_{i,n,\mu,s} E_n \c_{in\mu s}^\dag \c_{in\mu s},
  \label{eq:H_0}
\end{align}
where $\c_{in\mu s}$ is a fermionic operator which annihilates a
single atom on lattice site $i\in\mathbb{Z}^3$ in motional state
$n\in\mathbb{N}_0^3$ with nuclear spin $\mu\in\set{-I,-I+1,\cdots,I}$
(i.e.~projected onto a quantization axis) and orbital state
$s\in\set{\g,\e}$; and $E_n$ is the energy of a single atom in
motional state $n$.  In a harmonic trap approximation we would have
$E_n=\p{3/2+n_\x+n_\y+n_\z}\omega$ for an on-site angular trap
frequency $\omega$, but in general the aharmonicity of the lattice
potential will cause a non-negligible shift in motional state
energies.

In the absence of hyperfine coupling, as when addressing the spinless
${}^1S_0$ ($\g$) and ${}^3P_0$ ($\e$) orbital states of AEAs,
interactions between any two atoms are governed by their orbital
states alone, and are therefore characterized by four scattering
lengths $a_X$ with $X\in\set{\g\g,\e\g^-,\e\g^+,\e\e}$, where the $+$
($-$) superscript denotes symmetrization (anti-symmetrization) of a
two-body orbital state under particle exchange.  In the low-energy
limit, we can write the bare two-body interaction Hamiltonian in the
form \cite{gorshkov2010twoorbital}
\begin{align}
  H_{\t{int}} = \sum_{\substack{\mu<\nu\\s}}
  G_s \int \d^3x~ \hat\rho_{\mu s} \hat\rho_{\nu s}
  + G_+ \sum_{\mu,\nu}
  \int \d^3x~ \hat\rho_{\mu,\e} \hat\rho_{\nu,\g}
  + G_- \sum_{\mu,\nu} \int \d^3x~
  \hat\psi_{\mu,\e}^\dag \hat\psi_{\nu,\g}^\dag
  \hat\psi_{\nu,\e} \hat\psi_{\mu,\g},
  \label{eq:H_int_start}
\end{align}
where $\hat\psi_{\mu s}$ is a fermionic field operator for atoms with
nuclear spin $\mu$ and orbital state $s$;
$\hat\rho_{\mu s}\equiv\hat\psi_{\mu s}^\dag\hat\psi_{\mu s}$ is an
atomic density field operator; and the coupling constants $G_X$ are
defined in terms of the scattering lengths by
\begin{align}
  G_{s=\g,\e} \equiv \f{4\pi}{\mA}~ a_{ss}, &&
  G_\pm \equiv \f{2\pi}{\mA}~ \p{a_{\e\g+} \pm a_{\e\g-}},
  \label{eq:couplings}
\end{align}
where $\mA$ is the mass of a single atom.  Defining for brevity
\begin{align}
  G^{qr}_{st} \equiv \left\{
    \begin{array}{ll}
      G_q & ~ q = r = s = t \\
      G_+ & ~ q \ne r ~ \t{and} ~ (q,r) = (s,t) \\
      G_- & ~ q \ne r ~ \t{and} ~ (q,r) = (t,s) \\
      0 & ~ \t{otherwise}
    \end{array}\right.,
  \label{eq:coupling_tensor}
\end{align}
where $q,r,s,t\in\set{\g,\e}$ are orbital state indices, we can
alternately write the bare two-body interaction Hamiltonian in the
more compact form
\begin{align}
  H_{\t{int}}
  = \f12 \sum_{\substack{q,r,s,t\\\mu,\nu}} G^{qr}_{st} \int \d^3x~
  \hat\psi_{\mu s}^\dag \hat\psi_{\nu t}^\dag
  \hat\psi_{\nu r} \hat\psi_{\mu q}.
  \label{eq:H_int_fields}
\end{align}
For nuclear spins $\mu,\nu$, the symbol $G^{qr}_{st}$ gives the
coupling constant between the two-atom states
$(\mu,q)+(\nu,r)\leftrightarrow(\mu,s)+(\nu,t)$.

@@

  = -G^{qr}_{ts} \c_{\nu s}^\dag \c_{\mu t}^\dag \c_{\mu r} \c_{\nu q}

@@

In addition to the states in \eqref{eq:states_S}, the multi-body
Hamiltonian $H_M$ in \eqref{eq:H_M} has an $\p{N-1}$-fold degenerate
excited-state eigenspace which is asymmetric in the orbital degrees of
freedom, spanned by the states
\begin{align}
  \ket{\N,-_j} \equiv \f1{\sqrt2}
  \p{\c_{\mu_1,\e}^\dag \c_{\mu_1,\g} - \c_{\mu_j,\e}^\dag \c_{\mu_j,\g}}
  \ket{\N,0}
  \label{eq:states_A}
\end{align}
for $j=2,\cdots,N$.  If $N>2$, the asymmetric states are not separable
in their orbital and nuclear spin degrees of freedom.  An important
feature of the excited states in \eqref{eq:states_S} and
\eqref{eq:states_A} is that they are entirely independent of $M$,
which implies that the effect of multi-body interactions is simply to
modify the many-body atomic energy spectra without affecting the
energy eigenstates.  The eigenvalues $E_{NX}^{(M)}=\bk{\N X|H_M|\N X}$
of $H_M$ associated with each of the eigenstates in
\eqref{eq:states_S} and \eqref{eq:states_A} are provided in table
\ref{tab:eigen}, both in terms of the coefficients $U_{MX}$ of $H_M$
as expressed in \eqref{eq:H_M} and the $M$-body eigenvalues
$E_{MX}^{(M)}$.  Due to the SU($N$) symmetry of the multi-body
Hamiltonian $H_M$, the eigenvalues $E_{NX}^{(M)}$ depend on the number
of nuclear spins on a lattice site, $N$, but not on the actual nuclear
spins $\mu\in\N$ which are occupied.  The total $N$-body interaction
energies $E_{NX}$ are given in terms of the $M$-body eigenvalues
$E_{NX}^{(M)}$ by $E_{NX}=\sum_ME_{NX}^{(M)}$.

>>>>

% "left vector" arrow; requires tikz package
\usepackage{tikz}
\newcommand{\lvec}[1]
{\reflectbox{\ensuremath{\vec{\reflectbox{\ensuremath{#1}}}}}}

@@

Although an external trapping potential will generally break discrete
translational symmetry of the lattice, any background inhomogeneity
can be made negligible by spectroscopically addressing a sufficiently
small region of the lattice \cite{goban2018emergence}.  Throughout
this paper, we work strictly in the deep-lattice regime with
negligible tunneling between lattice sites.  We also neglect any
lattice inhomogeneities and assume that both atomic orbital states
(i.e.~${}^1S_0$ and ${}^3P_0$) experience identical lattice
potentials.  The single-particle Hamiltonian of the atoms can then be
written in the form
\begin{align}
  H_0 = \sum_{i,n,\mu,s} E_n \c_{in\mu s}^\dag \c_{in\mu s},
  \label{eq:H_0}
\end{align}
where $\c_{in\mu s}$ is a fermionic operator which annihilates a
single atom on lattice site $i\in\mathbb{Z}^3$ in motional state
$n\in\mathbb{N}_0^3$ with nuclear spin $\mu\in\set{-I,-I+1,\cdots,I}$
(i.e.~projected onto a quantization axis) and orbital state
$s\in\set{\g,\e}$; and $E_n$ is the energy of a single atom in
motional state $n$.  In a harmonic trap approximation we would have
$E_n=\p{3/2+n_\x+n_\y+n_\z}\omega$ for an on-site angular trap
frequency $\omega$, but in general the aharmonicity of the lattice
potential will cause a non-negligible shift in motional state
energies.

In the absence of hyperfine coupling, as when addressing the spinless
${}^1S_0$ ($\g$) and ${}^3P_0$ ($\e$) orbital states of AEAs,
interactions between any two atoms are governed by their orbital
states alone, and are therefore characterized by four scattering
lengths $a_X$ with $X\in\set{\g\g,\e\g^-,\e\g^+,\e\e}$, where the $+$
($-$) superscript denotes symmetrization (anti-symmetrization) of a
two-body orbital state under particle exchange.  In the low-energy
limit, we can write the bare two-body interaction Hamiltonian in the
form \cite{gorshkov2010twoorbital}
\begin{align}
  H_{\t{int}} = \sum_{\substack{\mu<\nu\\s}}
  G_s \int \d^3x~ \hat\rho_{\mu s} \hat\rho_{\nu s}
  + G_+ \sum_{\mu,\nu}
  \int \d^3x~ \hat\rho_{\mu,\e} \hat\rho_{\nu,\g}
  + G_- \sum_{\mu,\nu} \int \d^3x~
  \hat\psi_{\mu,\e}^\dag \hat\psi_{\nu,\g}^\dag
  \hat\psi_{\nu,\e} \hat\psi_{\mu,\g},
  \label{eq:H_int_start}
\end{align}
where $\hat\psi_{\mu s}$ is a fermionic field operator for atoms with
nuclear spin $\mu$ and orbital state $s$;
$\hat\rho_{\mu s}\equiv\hat\psi_{\mu s}^\dag\hat\psi_{\mu s}$ is an
atomic density field operator; and the coupling constants $G_X$ are
defined in terms of the scattering lengths $a_Y$ by
\begin{align}
  G_{s=\g,\e} \equiv \f{4\pi}{\mA}~ a_{ss}, &&
  G_\pm \equiv \f{2\pi}{\mA}~ \p{a_{\e\g+} \pm a_{\e\g-}},
  \label{eq:couplings}
\end{align}
where $\mA$ is the mass of a single atom.  Defining for brevity
\begin{align}
  G^{qr}_{st} \equiv \left\{
    \begin{array}{ll}
      G_q & ~ q = r = s = t \\
      G_+ & ~ q \ne r ~ \t{and} ~ (q,r) = (s,t) \\
      G_- & ~ q \ne r ~ \t{and} ~ (q,r) = (t,s) \\
      0 & ~ \t{otherwise}
    \end{array}\right.,
  \label{eq:coupling_tensor}
\end{align}
where $q,r,s,t\in\set{\g,\e}$ are orbital state indices, we can
alternately write the bare two-body interaction Hamiltonian in the
more compact form
\begin{align}
  H_{\t{int}}
  = \f12 \sum_{\substack{q,r,s,t\\\mu,\nu}} G^{qr}_{st} \int \d^3x~
  \hat\psi_{\mu s}^\dag \hat\psi_{\nu t}^\dag
  \hat\psi_{\nu r} \hat\psi_{\mu q}.
  \label{eq:H_int_fields}
\end{align}
For nuclear spins $\mu,\nu$, the symbol $G^{qr}_{st}$ gives the
coupling constant between the two-atom states
$(\mu,q)+(\nu,r)\leftrightarrow(\mu,s)+(\nu,t)$.

@@

  = -G^{qr}_{ts} \c_{\nu s}^\dag \c_{\mu t}^\dag \c_{\nu r} \c_{\mu q}

@@

In addition to the states in \eqref{eq:states_S}, the multi-body
Hamiltonian $H_M$ in \eqref{eq:H_M} has an $\p{N-1}$-fold degenerate
excited-state eigenspace which is asymmetric in the orbital degrees of
freedom, spanned by the states
\begin{align}
  \ket{\N,-,j} \equiv \f1{\sqrt2}
  \p{\c_{\mu_1,\e}^\dag \c_{\mu_1,\g} - \c_{\mu_j,\e}^\dag \c_{\mu_j,\g}}
  \ket{\N,0}
  \label{eq:states_A}
\end{align}
for $j=2,\cdots,N$.  If $N>2$, the asymmetric states are not separable
in their orbital and nuclear spin degrees of freedom.  An important
feature of the excited states in \eqref{eq:states_S} and
\eqref{eq:states_A} is that they are entirely independent of $M$,
which implies that the effect of multi-body interactions is simply to
modify the many-body atomic energy spectra without affecting the
energy eigenstates.  The eigenvalues $E_{NX}^{(M)}=\bk{\N X|H_M|\N X}$
of $H_M$ associated with each of the eigenstates in
\eqref{eq:states_S} and \eqref{eq:states_A} are provided in table
\ref{tab:eigen}, both in terms of the coefficients $U_{MX}$ of $H_M$
as expressed in \eqref{eq:H_M} and the $M$-body eigenvalues
$E_{MX}^{(M)}$.  Due to the SU($N$) symmetry of the multi-body
Hamiltonian $H_M$, the eigenvalues $E_{NX}^{(M)}$ depend on the number
of nuclear spins on a lattice site, $N$, but not on the actual nuclear
spins $\mu\in\N$ which are occupied.  The total $N$-body interaction
energies $E_{NX}$ are given in terms of the $M$-body eigenvalues
$E_{NX}^{(M)}$ by $E_{NX}=\sum_ME_{NX}^{(M)}$.
